---
title: "Tecan exploration and growth curve fitting"
author: "Chacon"
date: "2022-11-07"
output: rmarkdown::github_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
rm(list = ls())
source("./tecan_funcs.r")

```

## Tecan data

This shows how to use some helper functions to read in TECAN data and calculate growth rates. See the other .md, "different_growth_models.md," to better understand the different growth models one can use to fit.

Included here is OD data from a TECAN read. To get this, we save the OD "chunk" of data from the xls that the TECAN generates. Open the .csv in xcel if you wish to see what this file should look like.

Here is loading. Note that we specify that the response variable here is "OD" (measure = "OD"). If we were also loading in CFP or other fluorescence data, we'd want to change that argument. 

The gist of read_tecan_csv is to get the data into long-format, which is what we want for plotting and curve fitting.

```{r cars}
OD_path = "./25oct16_OD.csv"
OD = read_tecan_csv(OD_path, sep = ",", measure = "OD")
head(OD)
```

If we were loading multiple data sources (OD, CFP, YFP), we'd want to merge them all together. To do this, we MUST remove the seconds, hour, and temp from the subsequent data sources, as these are not an exact match due to fluctuations between measurements. Below is an example of how we'd do that, though it is faking having different data and actually just reloading the same data twice.

```{r combinedata}
OD_path = "./25oct16_OD.csv"
OD = read_tecan_csv(OD_path, sep = ",", measure = "OD")
fake_CFP_path = "./25oct16_OD.csv"
# notice how we select -hour, seconds, temp to remove these
CFP = read_tecan_csv(OD_path, sep = ",", measure = "CFP") %>%
  select(-c(hour, seconds, temp))
all_data = OD %>% left_join(CFP)
head(all_data)

```

## Plotting

I included a couple of helper functions to plot the plate in the same shape as the actual plate. It returns the ggplot object, so you can style it however you wish after the fact.

```{r plotplate}
fig = plot_plate(OD, "OD")
fig
fig + theme_bw(12) + scale_y_log10() + labs(title = "semi-log plot")
```

## metadata

Previously I had a function which would read a plate-shaped .csv and use this to add metadata (e.g. treatments). I've since realized that I can't predict everyone's usage, so instead here I'll show the most general (albeit cumbersome) way to add metadata. Then, I use plot_plate with a color argument to plot the metadata. 

```{r metadata}
# note: these aren't the actual treatments from that plate, its just a demo
OD$media = "water" # everything not specified gets this treatment
OD$media[OD$well %in% c("B2","B3","B4","B5")] = "succinate"
OD$media[OD$well %in% c("C7","C8","C9","C10")] = "glucose"

# then we can plot the treatment colors too
plot_plate(OD, measure = "OD", color = "media") + scale_y_log10()


```

## Subtracting blanks is critical 

One of the main things people wish to do is measure growth rates. See the other .md for a better understanding of the different growth functions. 

Critically, one needs to have the data start at a reasonable y-value, because almost all of the fitting functions depend on a reasonably y0 from which exponential (or logistic) growth can progress. 

Ideally, one would subtract a blank, and then what is left is "correct." However, due to lower-limits on the TECAN's measurement ability, this may prove erroneous as well.

Here I do a compromise. I subtract the minimum value from each well. Then, under the assumption that we started with OD = 0.5 / 200 (i.e. an OD = 0.5 stock diluted 1/200), we set any value that is below that to that.

```{r set_min_data}
OD = OD %>% group_by(well) %>% 
  mutate(OD = OD - min(OD)) %>%
  mutate(OD = ifelse(OD <= 0.5/200, 0.5/200, OD)) %>%
  ungroup()

plot_plate(OD)+labs(title = "y starts around 0.5 / 200")
```


## Measuring growth rates. 

For each of the growth-curve-fitting functions, there is an associated "fit_all_[]" function which applies it to a whole plate. 

Please note that non-linear curve fitting sometimes fails to converge. To remedy this, try increasing "tries," or smoothing your data (an example data smoothing is below).

Here I show how to do all four functions. They return a data frame which has as many rows as # of wells * # of parameters in model. I.e. it loses the temporal data. They can be merged together after the fact, however.

Importantly, other metadata (e.g. treatment columns) can be retained, by specifying the groups in which growth rate should be measured. This should always include well (because you want to have one growth rate per well), but here, could also include media. 

There is also a function "plot_plate_growthrates" which plots the growth rates in the shape of the plate. I use it for one set of growth rates below.

You'll see that some wells are missing data--that is because curve-fitting failed on those wells. 

```{r fitgrowthmodels}
loglin_gr = fit_all_loglinear(OD, measure = "OD", groups = c("well", "media"), tries = 50)

logistic_gr = fit_all_logistic(OD, measure = "OD", groups = c("well", "media"), tries = 50)

baranyi_gr = fit_all_baranyi(OD, measure = "OD", groups = c("well", "media"), tries = 50)

loglin_lm_gr = fit_all_loglinear_lm(OD, measure = "OD", groups = c("well", "media"), surround = 5)

plot_plate_growthrates(loglin_gr, response = "r")

```

As shown in the other .md, these models don't always yield identical growth rates. That said, they do correlate:

```{r modelcorr}
loglin_gr %>%
  mutate(loglinear_gr = r) %>%
  select(well, loglinear_gr) %>%
  left_join(baranyi_gr %>%
              mutate(baranyi_gr = r) %>%
              select(well, baranyi_gr)) %>%
  ggplot(aes(x = loglinear_gr, y = baranyi_gr))+
  geom_point()+
  stat_smooth(method = "lm")+
  ylim(0,1)+
  xlim(0,0.4)

loglin_gr %>%
  mutate(loglinear_gr = r) %>%
  select(well, loglinear_gr) %>%
  left_join(loglin_lm_gr %>%
              mutate(loglin_lm_gr = r) %>%
              select(well, loglin_lm_gr)) %>%
  ggplot(aes(x = loglinear_gr, y = loglin_lm_gr))+
  geom_point()+
  stat_smooth(method = "lm")
```

Two things you could do to help with fitting are smoothing the data, or cap the data at the max value reached.

```{r capdata}
# smooth data with a median smoothing, window size 11
OD = OD %>%
  group_by(well) %>%
  arrange(hour) %>%
  mutate(OD_smooth = rollmedian(OD, 11, fill = NA))

# cap data at the maximum reached
OD = OD %>%
  group_by(well) %>%
  mutate(OD_capped = ifelse(cycle > cycle[OD == max(OD)][1], max(OD), OD))

plot_plate(OD, measure = "OD_smooth")+labs(title = "smoothed OD")
plot_plate(OD, measure = "OD_capped")+labs(title = "capped OD")
```

Finally, let's redo growth rates on the capped data, and then plot the predicted lines over this. We can do this for each curve-fitting model (but not the loglineaer_lm model, because that works differently). Here we'll do it just for loglinear and logistic.

To plot predicted lines, we join the original data with the growth rate dataframe, add a new column called "OD_pred" which is the fitted line, then use plot plate, and finally add the second set of information.

Note that making the OD_pred depends on using the growth model, which means we need to specify differently-named parameters for each curve.

```{r redogrowthrates}

loglin_gr = fit_all_loglinear(OD, measure = "OD_capped", groups = c("well", "media"), tries = 50)

logistic_gr = fit_all_logistic(OD, measure = "OD_capped", groups = c("well", "media"), tries = 50)

left_join(OD, loglin_gr) %>%
  rowwise() %>%
  mutate(OD_pred = log_linear(hour, r, y0,lag, end)) %>%
  ungroup %>%
  plot_plate(measure = "OD_pred")+ # here is plot plate
  geom_line(aes(y = OD), size = 2, color = "black")+
  geom_line(size = 1, color = "blue")+
  scale_y_log10()


left_join(OD, logistic_gr) %>%
  rowwise() %>%
  mutate(OD_pred = logistic(hour, r, lag,K, y0)) %>%
  ungroup %>%
  plot_plate(measure = "OD_pred")+ # here is plot plate
  geom_line(aes(y = OD), size = 2, color = "black")+
  geom_line(size = 1, color = "blue")+
  scale_y_log10()

```

## Max growth summary joined to growth rate summary

Finally, I included a function to get yields. This should obviously be done on blank-subtrated data! It optionally smooths the data with a rolling mean smoothing first, to remove spikes. Set this to 1 (or just don't set it) to not smooth first. 

You can include which grouping variables to group by, and then after the fact, it is straightforward to join this to a growth rate results. 

At the very end I have a commented line which saves one of the resulting dataframes. 

```{r maxgrowth}
results = get_yields(OD, measure = "OD", groups = c("well","media"), smooth = 5) %>%
  left_join(loglin_gr)

head(results %>% filter(media != "water"))

plot_plate_growthrates(results, response = "yield", color = "media")

results %>%
  ggplot(aes(x = r, y = yield, color = media))+
  geom_point()

results %>%
  ggplot(aes(x = lag, y = r, color = media))+
  geom_point()

# how to write to file:
# write_csv(results, "growthrate_and_yield.csv")

```